---
title: "Time Series Analysis Homework08 Explaination"
author: "Hung-Ta, Wang (R13546017)"
date: "December 08, 2024"
output: pdf_document
latex_engine: xelatex
---

## 2. Recall the dataset “robot” firstly introduced in TSA HW06.

### 2.a. Use IMA(1, 1) to forecast five values ahead and calculate the 95% confidence intervals.

```{r message=FALSE, warning=FALSE}
library(forecast)
library(TSA)
library(tseries)

robot_data <- read.csv("C:/Git_Code/Some-practice/TSA HW06.robot.csv")
robot_ts <- ts(robot_data$robot)
robot_ima <- Arima(robot_ts, order = c(0, 1, 1))
forecast_result_ima <- forecast(robot_ima, h = 5)
print(forecast_result_ima)
```

#### Explanation:

We can see that all forecast values are in 95% confidence intervals.

### 2.b. Display the actual values, the five forecasts and the 95% confidence intervals of the five forecasts, all in

```{r message=FALSE, warning=FALSE}
plot(forecast_result_ima, 
     main = "Forecast with 95% Confidence Intervals of IMA(1,1)", 
     xlab = "t (index)", ylab = "Values")
```

#### Explanation:

We can see that the blue line is in the gray zone.

### 2.c. Use ARMA(1, 1) to forecast five values ahead and calculate the 95% confidence intervals. Compare the results with those in (a), what do you observe?

```{r message=FALSE, warning=FALSE}
robot_arma <- Arima(robot_ts, order = c(1, 0, 1))
forecast_result_arma <- forecast(robot_arma, h = 5)
print(forecast_result_arma)
plot(forecast_result_arma, 
     main = "Forecast with 95% Confidence Intervals of ARMA(1,1)", 
     xlab = "t (index)", ylab = "Values")
```

#### Explanation:

We can see that all forecast values are in 95% confidence intervals just like the blue line is in the gray zone.


## 3. The dataset “boardings” contains the monthly number of passengers who boarded light rail trains and buses in Denver, Colorado, from August 2000 to March 2006.

### 3.a. Plot the time series and tell your observation if there exists seasonality and if the series is stationary. 

```{r message=FALSE, warning=FALSE}
boardings_data <- read.csv("C:/Git_Code/Some-practice/TSA HW08.boardings.csv")
boardings_ts <- ts(boardings_data$log_boardings, start = c(2000, 8), frequency = 12)
plot(boardings_ts, main = "Time Series of Boardings", 
     xlab = "t (index)", ylab = "Log Boardings")
decomposed_boardings_ts <- decompose(boardings_ts)
plot(decomposed_boardings_ts)
```

#### Explanation:

We can see that the observation exists seasonality by the graph of decomposed time series. $\\$
The series is not stationary by its seasonality and trend.

### 3.b. Plot the sample ACF and see what are the significant lags?

```{r message=FALSE, warning=FALSE}
acf(boardings_ts, main = "Sample ACF of Time Series of Boardings")
```

#### Explanation:

We can observe that the $1^{st}, 5^{th}, 6^{th}, 7^{th}, 12^{th}$ bar is out of the interval. These are the significant lags.

### 3.c. Fit the data with ARMA(0, 3) × (1,0)_12, evaluate if the estimated coefficients $\{ \theta_1, \theta_2, \theta_3, \phi_{12} \}$ are significant. 

```{r message=FALSE, warning=FALSE}
pacf(boardings_ts, main = "Sample PACF of Time Series of Boardings")


model <- Arima(boardings_ts, order = c(0, 0, 3), seasonal = c(1, 0, 0))
summary(model)
coefficients <- coef(model)
standard_errors <- sqrt(diag(model$var.coef))
t_values <- coefficients / standard_errors
results <- data.frame(
  Coefficient = coefficients,
  `Standard Error` = standard_errors,
  `t-value` = t_values
)
print(results)
```

#### Explanation:

We have degree of freedom: 69-1 = 68. If t.value is greater than 2 or less than 2, it is significant.
For 95% confidence interval, we can see that t.value are all greater than 2. Thus, all coefficients are significant.
However, t.value of ma1, ma2, ma3 are decreasing since ma1 is significant by observing ACF.

## 4. The monthly airline passengers, first investigated by Box and Jenkins in 1976, is considered as the classic time series dataset (see “TSA HW08.airpass.csv”).

### 4.a. Plot the time series in its original scale and the log-transformed scale. Do you think making the log-transformation is appropriate?

```{r message=FALSE, warning=FALSE}
airpass_data <- read.csv("C:/Git_Code/Some-practice/TSA HW08.airpass.csv")
airpass_ts <- ts(airpass_data$airpass, start = c(1976,1) , frequency = 12)
log_airpass_ts <- log(airpass_ts)

plot(airpass_ts, 
     main = "Time Series of Airpass", 
     xlab = "t (index)", ylab = "Passengers")
plot(log_airpass_ts, 
     main = "Time Series of Airpass(log)", 
     xlab = "t (index)", ylab = "Passengers(log)")
```

#### Explanation:

We can see that there are trend and seasonality in the time series no matter it is log-transformed. From the aspect of making data stationary, I do not think log-transformation is an appropriate way.

### 4.b. Make the first-order difference over the “log-transformed” data. What do you observe?

```{r message=FALSE, warning=FALSE}
diff_log_airpass_ts <- diff(log_airpass_ts)
plot(diff_log_airpass_ts, 
     main = "Time Series of Airpass(log and diff)", 
     xlab = "t (index)", ylab = "Passengers(log and diff)")


acf(log_airpass_ts, main = "ACF of log_airpass_ts")
acf(diff_log_airpass_ts, main = "ACF of diff_log_airpass_ts")
```

#### Explanation:

We can see that the first-order difference over the log-transformed data may be not stationary. However, from the aspect of stationarity , it is much better than log-transformation.

### 4.c. Make a seasonal difference of the resulted series in (b), what do you observe?

```{r message=FALSE, warning=FALSE}
seasonality_diff_log_airpass_ts <- diff(diff_log_airpass_ts, lag = 12)
plot(seasonality_diff_log_airpass_ts, 
     main = "Time Series of Airpass(log and diff and seasonality)", 
     xlab = "t (index)", ylab = "Passengers(log and diff and seasonality)")
```

#### Explanation:

We can see that the time series seems to be more stationary than the resulted series in (b). However, we still cannot ensure that it is stationary merely by the graph. 

### 4.d. Plot the sample ACF of the resulted series in (c), explain what you see.

```{r message=FALSE, warning=FALSE}
acf(seasonality_diff_log_airpass_ts, 
    main = "Sample ACF of Time Series of Airpass(log and diff and seasonality)")
```

#### Explanation:

We can see that there are more sample ACF in the confidence interval which means there are less ACF are significant. However, we cannot observe an obvious pattern, such as cut-off in the ACF graph. However, from the aspect of stationarity , it is a bit better than the resulted series in (b).

### 4.e. Fit an ARIMA(0,1,1) × (0,1,1)_12 model to the log-transformed series. Diagnose the residuals of this model, including the sample ACF and the normality test

```{r message=FALSE, warning=FALSE}
airpass_sarima <- Arima(log_airpass_ts, order = c(0, 1, 1), seasonal = c(0, 1, 1))
summary(airpass_sarima)
residuals <- residuals(airpass_sarima)
checkresiduals(residuals)
acf(residuals, main = "Sample ACF of Residuals")
adf.test(residuals)
qqnorm(residuals)
qqline(residuals)
```

#### Explanation:

Sample ACF shows that the ACF of the residuals are separate and there is not a obvious pattern. We guess that the ACF may be stationary.
Moreover, We can see that the Ljung Box test and ADF test are all show that the residuals is a stationary process.
From Normal Q Q Plot, The residuals are approximately normal, it may slightly not follow normal distribution on two side.

### 4.f. Make forecasts for “two” years based on the model in (e). The confidence intervals shall be included.

```{r message=FALSE, warning=FALSE}
airpass_sarima_seasonality_diff_log_airpass_ts <- Arima(seasonality_diff_log_airpass_ts, order = c(0, 1, 1), seasonal = c(0, 1, 1))
forecasts_airpass_sarima_seasonality_diff_log_airpass_ts <- forecast(airpass_sarima_seasonality_diff_log_airpass_ts, h = 24)
plot(forecasts_airpass_sarima_seasonality_diff_log_airpass_ts, main = "Two-Year Forecast with Confidence Intervals",
     xlab = "Time", ylab = "airpass_sarima_seasonality_diff_log_airpass_ts")

airpass_sarima_airpass_ts <- Arima(airpass_ts, order = c(0, 1, 1), seasonal = c(0, 1, 1))
forecasts_Passengers <- forecast(airpass_sarima_airpass_ts, h = 24)
plot(forecasts_Passengers, main = "Two-Year Forecast with Confidence Intervals",
     xlab = "Time", ylab = "Air Passengers(Original)")
print(forecasts_airpass_sarima_seasonality_diff_log_airpass_ts)
print(forecasts_Passengers)
```

#### Explanation:

We can see that no matter original and seasonality diff log data are all in confidence interval.
